---
title: "Homework 2 Report"
output: html_document
---


## Name

Shaikh Ibrahim Noman

### Setup

Read in the data with `read_csv()` and store the data as an R object named `dataset`. Check the data to make sure all of the expected observations and variables are there.


```{r}
## Load the data and any necessary packages here.
library(tidyverse)
dataset <- read.csv('maacs.csv.gz')
head(dataset, 10)
```

## Part 1

We will first consider the relationship between FEV1 and age. In general, it is expected that as children get older (and hence, larger in size), their FEV1 values should get higher. 

Consider the statement "FEV1 values in children are higher in older children relative to younger children". 

Write a function in R that takes the `dataset` object as an argument and returns `TRUE` if the statement above is true for the dataset and `FALSE` otherwise. 

NOTE: In order to write this function, you will need to translate the statement above into something that can be checked with the data. There are many ways in which you can do that translation correctly and you only need to pick one way here. 

NOTE: For this part, do not use any plots.

```{r}
## To see the relationship between two variables you can check the correlation
fev1_age_relationship <- function(dataset){
  
  # Calculating the correlation between age and fev1
  correlation = cor(dataset$age, dataset$fev1, use = 'complete.obs')
  
  # Checking using if-else
  # If variables are directly correlated (Both go up together) return true
  if (correlation >= 0.75){
    return(TRUE)
  # Otherwise return false
  } else {
    return(FALSE)
  }
}

# Calling the function
fev1_age_relationship(data = dataset)
```


## Part 2

Fit a linear regression model with FEV1 as the outcome and age as a predictor.

How much does FEV1 change for a 1-year increase in the child's age? 

FEV1 increases by 0.171 liters for each 1-year increase in the child's age.

```{r}
model <- lm(fev1 ~ age,  data = dataset)
summary(model)
```


Write your data analysis statement interpreting the regression model here:

Since the p-value for age is less than 0.05, age is a statistically significant predictor of FEV1. The positive regression coefficient indicates that FEV1 tends to increase as age increases.

## Part 3

Develop **three** supporting premises derived from the data that support the statement you wrote in Part 2. These can be plots, other summary statistics, or model results.

NOTE: At least one supporting premise should use a plot.

```{r}
# 1. Scatter plot
ggplot(dataset) +
  aes(x = age, y = fev1) +
  geom_point(color = 'red')

# 2. Getting mean in different age groups (summary statistics)
## Create two age groups: younger and older
age_data <- dataset %>%
  mutate(age_group = ifelse(age <= median(age, na.rm = TRUE), 'Younger', 'Older'))

## Compare mean FEV1 between groups
group_means <- age_data %>%
  group_by(age_group) %>%
  summarise(mean_fev1 = mean(fev1, na.rm = TRUE))
group_means

# 3. R-Squared (Model Output)
model_summary <- summary(model)
r_squared <- model_summary$r.squared
cat('R-Squared:', r_squared)
```


Write the three supporting premise statements here:

1. In the scatter plot, there is a direct positive relationship between `age` and `fev1`.

2. When analyzing the mean of `fev1` between older and younger children, we can see the mean is greater for older (2.54) compared to younger (1.33)

3. The regression model that includes age explains about 64.7% of the variation in fev1 (R-squared = 0.6469). This relatively high R-squared indicates that age is an important predictor and provides strong evidence of its positive association with fev1.


## Part 4

For each of the supporting premises above, write a function that takes the `dataset` object as an argument and returns `TRUE` if the supporting premise statement above is true for the dataset and `FALSE` otherwise. 

For statements involving plots, instead of returning `TRUE` or `FALSE`, you function should do two things:

1. Produce the plot that is used in the statement

2. Produce a hypothetical version of the plot in the event that the statement is true. This can be done using simulated data or by simply hand drawing a plot.


```{r}
## Function for supporting premise statement 1
premise1_plot <- function(dataset){
  
  # 1. Producing the statement plot
  statement_plot <- ggplot(dataset) +
    aes(x = age, y = fev1) +
    geom_point(color = 'red')
  
  print(statement_plot)
  
  # 2. Simulate data assuming statement is true
  set.seed(7) # for reproducibility
  n <- nrow(dataset)
  hypothetical_data <- data.frame(
    age = dataset$age,
    fev1 = 1 + 0.1 * dataset$age + rnorm(n, 0, 0.2) # linear increase with small noise
  )
  
  hypothetical_plot <- ggplot(hypothetical_data) +
    aes(x = age, y = fev1) +
    geom_point(color = 'blue') 
  
  print(hypothetical_plot)
}

premise1_plot(dataset = dataset)
```


```{r}
## Function for supporting premise statement 2
premise2 <- function(dataset){
  
  # Calculating the group means
  age_data <- dataset %>%
  mutate(age_group = ifelse(age <= median(age, na.rm = TRUE), 'Younger', 'Older'))
  group_means <- age_data %>%
    group_by(age_group) %>%
    summarise(mean_fev1 = mean(fev1, na.rm = TRUE))
  
  # Extract means
  younger_mean <- group_means$mean_fev1[group_means$age_group == 'Younger']
  older_mean   <- group_means$mean_fev1[group_means$age_group == 'Older']
  
  # If older mean > younger mean, return TRUE
  if (older_mean > younger_mean){
    return(TRUE)
  } else {
    return(FALSE)
  }
}

premise2(dataset = dataset)
```


```{r}
## Function for supporting premise statement 3
premise3 <- function(dataset){
  
  # Getting r-squared value
  model_summary <- summary(model)
  r_squared <- model_summary$r.squared
  
  # If else statement
  if (r_squared > 0.5){
    return(TRUE)
  } else {
    return(FALSE)
  }
}

premise3(dataset = dataset)
```

Execute each of your function and show that the produce the expected output.




## Part 5

Describe one alternative to the primary statement "FEV1 values in children are higher in older children relative to younger children". 

Alternative Statement: FEV1 values in children do not increase with age, and younger children may have similar or even higher FEV1 values compared to older children.

Fault tree is located in Canvas.


